submission

### 2024-03-04 submission.Rmd
feat: m-step + evaluation

  return(list(result = result, mixture_EM = mixture_EM, log_likelihood = log_likelihood))
}
```

@@

@@

```{r}
# Define log-likelihood function for Student's t-distribution
log_likelihood_student_t <- function(theta, x) {
  nu <- theta[1]
  mu <- theta[2]
  sigma <- theta[3]

@@

  init_params <- c(5, mean(y), sd(y))
  opt_result <- optim(init_params, log_likelihood_student_t, x = y, method = "L-BFGS-B",
                      lower = c(0, -Inf, 0), upper = c(Inf, Inf, Inf))

@@

# Plot histogram of the data and fitted PDF
ggplot() +
  geom_histogram(data = data.frame(y), aes(x = y, y = ..density..), bins = 30, fill = "skyblue", color = "black") +
  geom_line(data = plot_data, aes(x = x, y = pdf), color = "red", size = 1) +
  labs(title = "Fitted Student's t-distribution Model",
       x = "Daily Return",
       y = "Density") +
  theme_minimal()
```

@@

>>>>

  return(list(result = result, mixture_EM = mixture_EM, log_likelihood = log_likelihood, deviance= ))
}
```

@@

For the M-step we follow a similar procedure but now we have to change the update for the means. Which will take the following form. First we have the same parameters,

$$
\begin{align}
\hat{\pi}_k = \dfrac{\sum_{i=1}^K \tau_{ik}}{n} \hspace{5mm} \mu_k = \dfrac{\sum_{i=1}^K \tau_{ik} x_i}{\sum_{i=1}^K \tau_{ik}} \hspace{5mm} \hat{\sigma}_k^2 = \dfrac{\sum_{i=1}^K \tau_{ik} z_i^2}{\sum_{i=1}^K \tau_{ik}} - \mu_k^2
\end{align}
$$

Note that we do not denote the mean with $\hat{\mu}$ above as it is only used on the optimization for the variance, to compute the new $\hat{\mu}$ we do the following.

$$
\hat{\mu}_k = \frac{\sum_{i=1}^K w_k \mu_k}{\sum_{i=1}^K w_k}
$$

Where $w_k$ is computed by.

$$
w_k = \dfrac{\sum_{i=1}^K \tau_{ik}}{\hat{\sigma}_{k}^2}
$$

We can then use this to create a modified version of the EM algorithm to deterministically obtain components with equal means.

@@

```{r}
# Define log-likelihood function for Student's t-distribution
neglog_likelihood_student_t <- function(theta, x) {
  nu <- theta[1]
  mu <- theta[2]
  sigma <- theta[3]

@@

  init_params <- c(5, mean(y), sd(y))
  opt_result <- optim(init_params, neglog_likelihood_student_t, x = y, method = "L-BFGS-B",
                      lower = c(0, -Inf, 0), upper = c(Inf, Inf, Inf))

@@

# Plot histogram of the data and fitted PDF
ggplot() +
  geom_histogram(data = data.frame(y), aes(x = y, y = after_stat(density)), bins = 30, fill = "skyblue", color = "black") +
  geom_line(data = plot_data, aes(x = x, y = pdf), color = "red", size = 1) +
  labs(title = "Fitted Student's t-distribution Model",
       x = "Daily Return",
       y = "Density") +
  theme_minimal()
```

@@

To effectively evaluate the models we can employ a few techniques, the most obvious is to apply the BIC metric to measure how well the models capture the distribution of the data.

```{r}
bic_benchmark <- function(x, out_model, n_free_params) {
  n <- length(x)
  bic_values = list()

  return(out_model$deviance + log(n) * n_free_params)
}
```

```{r}
mu_data = mean(y)
pi = c(.2, .2, .2, .2, .2)
means = c(mu_data, mu_data, mu_data, mu_data, mu_data)
variances = c(10, 7, 5, 3, 1)

theta0  <- list(pi=pi, mu=means, sigma=variances)
out_general <- mixture_gaussian1D(sp500_data$DailyReturn, theta0, M_step = M_step_general)

b_5unconstrained <- bic_benchmark(y, out_general, 3 * 5)

out_general_constrained <- mixture_gaussian1D(sp500_data$DailyReturn, theta0, M_step = M_step_same_means)

b_5constrained <- bic_benchmark(y, out_general_constrained, 3 * 5)

theta = student_t_params
b_tdist = 3 * log(length(y)) + 2 * neglog_likelihood_student_t(c(theta$nu, theta$mu, theta$sigma), y)

single_guass <- fitdistr(sp500_data$DailyReturn, "normal")
b_single_gauss <- BIC(single_guass)
```

```{r}
rbind(
  five_comp_gmm             = b_5unconstrained,
  five_comp_gmm_constrained = b_5constrained,
  student_tdistribution     = b_tdist,
  single_gaussian           = b_single_gauss
) %>% as.data.frame() %>% setNames("BIC") %>% rmarkdown::paged_table()
```

```{r}
ks.test(y, "pmixture", out_general_constrained$theta)
```

```{r}
ks.test(y, "pmixture", out_general$theta)
```

```{r}
ks.test(y, "pmixture", theta0)
```
